---
title: "Exploration"
author: "Andrew Lucas"
format: html
execute:
  echo: false
---

## Explore NHL Shot Data by Season

Dive into 19 seasons of NHL shot data (2007–2025). Select a season from the dropdown to see where shots originate on the ice for that year. The **heatmap** shows shot density — brighter areas mean more shots come from there. The **scatter points** show individual shot locations, colored by player position.

**How to use this page:**

- **Season dropdown** (top-left) — Switch between any season from 2007 to 2025, or select "All Seasons" to see everything at once.
- **Legend** — Click a position name in the legend to toggle it on or off.
- **Hover** — Hover over any point or heatmap cell for details.

::: {.callout-note}
Shot data is sourced from MoneyPuck and includes all shot attempts (shots on goal, missed shots, and blocked shots) as well as goals. Positions shown are Center, Defense, Left Wing, and Right Wing.
:::

```{r}
#| label: packages
#| message: false
#| warning: false
library(data.table)
library(hexbin)
library(plotly)
```

```{r}
#| label: load-data
#| message: false
#| warning: false
#| cache: true
seasons <- 2007:2025
files <- file.path("data", paste0("shots_", seasons, ".csv"))

shots_raw <- rbindlist(lapply(files, function(f) {
  fread(f, select = c("arenaAdjustedXCord", "arenaAdjustedYCord",
                       "playerPositionThatDidEvent", "season", "event"))
}))
```

```{r}
#| label: prepare-data
shots <- shots_raw[
  event %in% c("SHOT", "GOAL") &
  playerPositionThatDidEvent %in% c("C", "D", "L", "R")
]

shots[, position_label := fcase(
  playerPositionThatDidEvent == "C", "Center",
  playerPositionThatDidEvent == "D", "Defense",
  playerPositionThatDidEvent == "L", "Left Wing",
  playerPositionThatDidEvent == "R", "Right Wing"
)]
```

```{r}
#| label: shot-map
#| fig-width: 8
#| fig-height: 6.5
#| message: false
#| warning: false

pos_colors <- c("Center" = "#E74C3C", "Defense" = "#3498DB",
                 "Left Wing" = "#2ECC71", "Right Wing" = "#9B59B6")
pos_names <- c("Center", "Defense", "Left Wing", "Right Wing")

# Pre-compute hexbin + scatter per season
hex_season <- list()
scat_season <- list()

for (s in seasons) {
  d <- shots[season == s]
  d_hex <- list()
  d_scat <- list()
  for (i in seq_along(pos_names)) {
    dp <- d[position_label == pos_names[i]]
    if (nrow(dp) == 0) {
      d_hex[[i]] <- data.frame(x = numeric(0), y = numeric(0), count = numeric(0))
      d_scat[[i]] <- dp
    } else {
      y_max <- max(abs(dp$arenaAdjustedYCord), na.rm = TRUE) + 5
      hb <- hexbin(dp$arenaAdjustedXCord, dp$arenaAdjustedYCord,
                   xbnds = c(-100, 100), ybnds = c(-y_max, y_max),
                   xbins = 40, IDs = TRUE)
      d_hex[[i]] <- data.frame(x = hb@xcm, y = hb@ycm, count = hb@count)
      set.seed(42)
      d_scat[[i]] <- dp[sample(.N, max(1, floor(.N * 0.10)))]
    }
  }
  hex_season[[as.character(s)]] <- d_hex
  scat_season[[as.character(s)]] <- d_scat
}

# Global zmax for consistent color scale across seasons
all_counts <- unlist(lapply(hex_season, function(season) {
  unlist(lapply(season, function(h) h$count))
}))
global_cmax <- max(all_counts, na.rm = TRUE)

# Rink shapes
rink_shapes <- list(
  list(type = "rect", x0 = -100, x1 = 100, y0 = -42.5, y1 = 42.5,
       fillcolor = "rgba(219,232,245,1)", line = list(color = "#888", width = 2),
       layer = "below"),
  list(type = "line", x0 = 0, x1 = 0, y0 = -42.5, y1 = 42.5,
       line = list(color = "#C0392B", width = 2.5), layer = "below"),
  list(type = "line", x0 = 25, x1 = 25, y0 = -42.5, y1 = 42.5,
       line = list(color = "#2980B9", width = 2.5), layer = "below"),
  list(type = "line", x0 = -25, x1 = -25, y0 = -42.5, y1 = 42.5,
       line = list(color = "#2980B9", width = 2.5), layer = "below"),
  list(type = "line", x0 = 89, x1 = 89, y0 = -42.5, y1 = 42.5,
       line = list(color = "#C0392B", width = 1.5), layer = "below"),
  list(type = "line", x0 = -89, x1 = -89, y0 = -42.5, y1 = 42.5,
       line = list(color = "#C0392B", width = 1.5), layer = "below"),
  list(type = "circle", x0 = 83, x1 = 95, y0 = -6, y1 = 6,
       line = list(color = "#2980B9", width = 1.5), fillcolor = "rgba(41,128,185,0.08)",
       layer = "below"),
  list(type = "circle", x0 = -95, x1 = -83, y0 = -6, y1 = 6,
       line = list(color = "#2980B9", width = 1.5), fillcolor = "rgba(41,128,185,0.08)",
       layer = "below"),
  list(type = "circle", x0 = 54, x1 = 84, y0 = -37, y1 = -7,
       line = list(color = "#C0392B", width = 1.2, dash = "dot"),
       fillcolor = "rgba(0,0,0,0)", layer = "below"),
  list(type = "circle", x0 = 54, x1 = 84, y0 = 7, y1 = 37,
       line = list(color = "#C0392B", width = 1.2, dash = "dot"),
       fillcolor = "rgba(0,0,0,0)", layer = "below"),
  list(type = "circle", x0 = -84, x1 = -54, y0 = -37, y1 = -7,
       line = list(color = "#C0392B", width = 1.2, dash = "dot"),
       fillcolor = "rgba(0,0,0,0)", layer = "below"),
  list(type = "circle", x0 = -84, x1 = -54, y0 = 7, y1 = 37,
       line = list(color = "#C0392B", width = 1.2, dash = "dot"),
       fillcolor = "rgba(0,0,0,0)", layer = "below")
)

# Color scale for heatmap
hc <- list(list(0, "#FFFFCC"), list(0.15, "#FED976"),
           list(0.35, "#FD8D3C"), list(0.55, "#E31A1C"),
           list(0.75, "#BD0026"), list(1, "#800026"))

# Build all traces
p <- plot_ly(width = 800, height = 650)

for (s in seasons) {
  s_key <- as.character(s)
  hx <- hex_season[[s_key]]
  sc <- scat_season[[s_key]]

  for (i in 1:4) {
    p <- p %>% add_heatmap(
      x = hx[[i]]$x, y = hx[[i]]$y, z = hx[[i]]$count,
      colorscale = hc, zmin = 0, zmax = global_cmax,
      showscale = FALSE, showlegend = FALSE,
      hovertemplate = paste0(pos_names[i], ": %{z} shots<br>X: %{x:.0f} ft<br>Y: %{y:.0f} ft<extra></extra>")
    )
  }

  for (i in 1:4) {
    p <- p %>% add_markers(
      data = sc[[i]],
      x = ~arenaAdjustedXCord, y = ~arenaAdjustedYCord,
      marker = list(color = pos_colors[pos_names[i]], size = 5, opacity = 0.45,
                    line = list(color = "white", width = 0.3)),
      name = pos_names[i],
      hovertemplate = paste0(pos_names[i], "<br>X: %{x:.0f} ft<br>Y: %{y:.0f} ft<extra></extra>")
    )
  }
}

# Set initial visibility: only show the most recent season
p <- plotly_build(p)
most_recent <- as.character(max(seasons))
for (idx in seq_along(seasons)) {
  s_key <- as.character(seasons[idx])
  is_vis <- s_key == most_recent
  start <- (idx - 1) * 8 + 1
  end <- idx * 8
  for (t in start:end) {
    p$x$data[[t]]$visible <- is_vis
  }
}

# Season dropdown buttons
n_per_season <- 8
total_traces <- n_per_season * length(seasons)

# "All Seasons" button — show everything
all_vis <- as.list(rep(TRUE, total_traces))
all_button <- list(method = "restyle", label = "All Seasons",
                   args = list("visible", all_vis))

# Per-season buttons
season_buttons <- lapply(seq_along(seasons), function(idx) {
  vis <- rep(FALSE, total_traces)
  start <- (idx - 1) * n_per_season + 1
  end <- idx * n_per_season
  vis[start:end] <- TRUE
  list(method = "restyle", label = as.character(seasons[idx]),
       args = list("visible", as.list(vis)))
})

all_buttons <- c(list(all_button), season_buttons)

p <- p %>% layout(
  title = list(
    text = "NHL Shot Positions by Season",
    font = list(size = 18, family = "Arial"),
    x = 0.5, xanchor = "center"
  ),
  xaxis = list(title = "", range = c(-100, 100), zeroline = FALSE,
               showgrid = FALSE, scaleanchor = "y", scaleratio = 1,
               showticklabels = FALSE, fixedrange = TRUE),
  yaxis = list(title = "", range = c(-55, 55), zeroline = FALSE,
               showgrid = FALSE, showticklabels = FALSE, fixedrange = TRUE),
  legend = list(
    title = list(text = "Position", font = list(size = 12)),
    font = list(size = 11),
    bgcolor = "rgba(255,255,255,0.9)",
    bordercolor = "#ccc", borderwidth = 1,
    x = 1.02, y = 1, xanchor = "left"
  ),
  margin = list(t = 60, b = 40, l = 10, r = 110),
  plot_bgcolor = "rgba(219,232,245,1)",
  paper_bgcolor = "white",
  shapes = rink_shapes,
  updatemenus = list(
    list(
      type = "dropdown", x = 0.0, y = 1.12, xanchor = "left", yanchor = "top",
      bgcolor = "white", bordercolor = "#aaa", borderwidth = 1,
      font = list(size = 12),
      buttons = all_buttons
    )
  )
)

p
```
