---
title: "The 2012 Lockout"
author: "Andrew Lucas"
format: html
execute:
  echo: false
---

<div class="story-hook">
<h1>The season that almost wasn't.</h1>
<p>113 days of labor war. 510 games erased. A 48-game sprint from January to April.</p>
<p class="kicker">The 2012–13 NHL lockout, and what the shot data shows.</p>
</div>

---

<div class="story-section">

### What Happened

At 11:59 p.m. on September 15, 2012, the NHL's collective bargaining agreement expired — and the owners locked out the players. It was the third lockout under commissioner Gary Bettman, and the fight was over the same fault line as always: money. Players had been taking roughly **57% of hockey-related revenue** — a pool worth about $3.3 billion a year — while owners pushed for a 50–50 divide and a tighter salary cap.

The season was supposed to open on October 11. Instead, the calendar melted away. Preseason games canceled first, then regular-season games in waves, then the Winter Classic — the league's outdoor showcase — and the All-Star Weekend. In total, **510 regular-season games never happened**: 41.5% of the schedule.

Talks sputtered, broke down over the winter holiday, and nearly ended in court when the union considered disclaiming interest. Then, after a 16-hour bargaining session brokered by federal mediator Scot Beckenbaugh, a tentative deal landed at 4:45 a.m. on **January 6, 2013** — 113 days after the lockout began. The new CBA split revenue 50–50, limited contract lengths, and was ratified within the week.

</div>

```{r}
#| label: lockout-packages
#| echo: false
#| message: false
#| warning: false
library(data.table)
library(ggplot2)
```

```{r}
#| label: lockout-load-data
#| echo: false
#| message: false
#| warning: false
#| cache: true
s11 <- fread("data/shots_2011.csv",
             select = c("game_id", "event", "season",
                        "isPlayoffGame", "homeTeamCode"))
s12 <- fread("data/shots_2012.csv",
             select = c("game_id", "event", "season",
                        "isPlayoffGame", "homeTeamCode"))
```

```{r}
#| label: lockout-summary
#| echo: false
summarize_season <- function(d, label) {
  reg <- d[isPlayoffGame == 0]
  data.table(
    season_label = label,
    games        = uniqueN(reg$game_id),
    teams        = uniqueN(reg$homeTeamCode),
    shots        = nrow(reg[event %in% c("SHOT", "GOAL")]),
    goals        = nrow(reg[event == "GOAL"])
  )
}
season_summary <- rbind(
  summarize_season(s11, "2011-12"),
  summarize_season(s12, "2012-13")
)
season_summary[, shots_pg := shots / games]
season_summary[, goals_pg := goals / games]
```

```{r}
#| label: lockout-stat-grid
#| echo: false
#| output: asis
s_prev <- season_summary[season_label == "2011-12"]
s_lock <- season_summary[season_label == "2012-13"]
games_per_team <- round(2 * s_lock$games / s_lock$teams)
games_cancelled <- s_prev$games - s_lock$games
goals_pg <- sprintf("%.2f", s_lock$goals_pg)
goals_pg_prev <- sprintf("%.2f", s_prev$goals_pg)

cat(sprintf('
<div class="stat-grid">
<div class="stat-box"><div class="number">%d</div><div class="label">Games Per Team (vs. 82)</div></div>
<div class="stat-box"><div class="number">%d</div><div class="label">Games Never Played</div></div>
<div class="stat-box"><div class="number">%s</div><div class="label">Goals per Game &mdash; Unchanged from %s</div></div>
</div>
', games_per_team, games_cancelled, goals_pg, goals_pg_prev))
```

---

<div class="story-section">

### The Sprint

The deal resurrected a **48-game season** that ran from January 19 to April 28, 2013 — 99 days, roughly a game every other day. With no time to waste, the league scrapped all inter-conference games to cut travel and packed the schedule with back-to-backs after a one-week training camp.

The compressed chaos suited Chicago. The Blackhawks opened the season with 24 consecutive games without a regulation loss — an NHL record — and went on to win the Cup in June. The rest of the league simply tried to survive the treadmill.

</div>

---

<div class="story-section">

### By the Numbers

What did half a season do to the way the game was played? Almost nothing, at least per game. The shortened season produced **58.3 shots per game** versus **59.5** the year before — a 2% dip — and scoring was essentially frozen at **5.30 goals per game** in both seasons.

The lockout cut the *quantity* of hockey, not its character. Rate stats — the ones that describe how a game actually looks — shrugged off the disruption entirely.

</div>

```{=html}
<div class="section-label">
<h2 style="font-size:1.3rem; margin-bottom:0.5rem;">Pace Held, Volume Vanished</h2>
```

```{r}
#| label: lockout-plot-pace
#| echo: false
#| message: false
#| warning: false
#| fig-width: 8
#| fig-height: 4.6
plot_df <- melt(season_summary,
                id.vars = "season_label",
                measure.vars = c("shots_pg", "goals_pg"),
                variable.name = "metric",
                value.name = "value")
plot_df[, metric := factor(metric,
                           levels = c("shots_pg", "goals_pg"),
                           labels = c("Shots per Game", "Goals per Game"))]

ggplot(plot_df, aes(x = season_label, y = value, fill = season_label)) +
  geom_col(width = 0.55) +
  geom_text(aes(label = sprintf("%.1f", value)),
            vjust = -0.5, size = 4.5, fontface = "bold") +
  facet_wrap(~ metric, scales = "free_y", nrow = 1) +
  scale_fill_manual(values = c("2011-12" = "#3498DB", "2012-13" = "#C0392B")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(title = "Per-Game Pace Barely Flinched",
       subtitle = "Regular season, 2011-12 vs. lockout-shortened 2012-13",
       x = NULL, y = NULL) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold", hjust = 0.5, size = 15),
    plot.subtitle = element_text(hjust = 0.5, color = "grey40", size = 11),
    strip.text = element_text(face = "bold", size = 12),
    panel.grid.major.x = element_blank(),
    axis.text = element_text(size = 11)
  )
```

```{=html}
<p class="preview-caption">
  Shot and goal rates per game, computed from the MoneyPuck shot data.
</p>
</div>

```{=html}
<div class="section-label">
<h2 style="font-size:1.3rem; margin-bottom:0.5rem;">The Deepest Dip in 19 Seasons</h2>
```

```{r}
#| label: lockout-load-games-by-season
#| echo: false
#| message: false
#| warning: false
#| cache: true
games_by_season <- rbindlist(lapply(sprintf("data/shots_%d.csv", 2007:2025), function(f) {
  d <- fread(f, select = c("game_id", "isPlayoffGame", "season"))
  reg <- d[isPlayoffGame == 0]
  data.table(season = reg$season[1], games = uniqueN(reg$game_id))
}))
games_by_season[, highlight := season == 2012]
```

```{r}
#| label: lockout-plot-context
#| echo: false
#| message: false
#| warning: false
#| fig-width: 8
#| fig-height: 4.6
ggplot(games_by_season, aes(x = factor(season), y = games, fill = highlight)) +
  geom_col(width = 0.7) +
  geom_text(data = games_by_season[highlight == TRUE],
            aes(label = games), vjust = -0.5,
            size = 4.5, fontface = "bold", color = "#C0392B") +
  scale_fill_manual(values = c("TRUE" = "#C0392B", "FALSE" = "#AEB6BF"),
                    guide = "none") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  labs(title = "Regular-Season Games Played, 2007–2025",
       subtitle = "The lockout cost 510 games; only the COVID seasons come close",
       x = NULL, y = NULL) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, size = 15),
    plot.subtitle = element_text(hjust = 0.5, color = "grey40", size = 11),
    panel.grid.major.x = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
    axis.text.y = element_text(size = 10)
  )
```

```{=html}
<p class="preview-caption">
  Regular-season games per season from the shot data. The 2019-20 and 2020-21 dips are COVID-shortened seasons.
  <a href="exploration.qmd">Compare seasons interactively &rarr;</a>
</p>
</div>
```

---

<div class="story-section">

### The Takeaway

1. **The calendar shrank; the game didn't.** Shot and goal rates per game were nearly identical on either side of the lockout.
2. **Half the sample.** Any comparison involving 2012-13 rests on 720 games instead of 1,230 — read raw totals carefully.
3. **Only a pandemic matched it.** Across 19 seasons, only the COVID interruptions produced a comparable hole in the schedule.

The ice didn't change when the lawyers left the room. Hockey, it turns out, is remarkably resistant to interruptions — as long as there are still games to play.

</div>

<div class="cta-box">
<h3>See it for yourself.</h3>
<p>Head to the <a href="exploration.qmd" style="font-weight:600;">Exploration page</a> and select season 2012 to explore the shortened year.</p>
</div>
